---
author: Kamal Rautela
output:
  md_document
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

R Markdown file for Course Reproducible Research Week2 Project 
===============================================================


[A] Loading and Preprocessing the data
========================================

```{r results="hide"}

rm(list=ls())
library(dplyr)
library(ggplot2)
activityData<-read.csv("activity.csv")

###Rename the variable names on the data frame to something more meaningful
activityData<-setNames(activityData,c("Steps","MeasureDate","Interval"))
###Ignore the record with "NA" for Steps
subActivityData<-subset(activityData,complete.cases(activityData))

```


[B] What is the mean total number of steps taken per day?
=====================================================


```{r}
###Group the data by date for each day in the dataset
grpd<-group_by(subActivityData,MeasureDate)

##B1. Calculate total number of steps taken each day
stepsPerDay<-data.frame(tapply(grpd$Steps,grpd$MeasureDate,sum,na.rm=TRUE))
stepsPerDay<-setNames(stepsPerDay,"Steps")
```

B2.Histogram of total number of steps taken each day 

```{r}
hist(stepsPerDay$Steps,xlab="Steps Per Day",main="Frequency of Steps per Day")

##B3.1 Calculate and report the mean number of steps take each day
meanStepsPerDay<-mean(stepsPerDay$Steps,na.rm = TRUE)

##B3.2 Calculate and report the median number of steps take each day
medianStepsPerDay<-median(stepsPerDay$Steps,na.rm = TRUE)

```

Steps per day:

B3.1 Mean of all the Steps per Day is: `r meanStepsPerDay`

B3.2 Median of all the Steps per Day is: `r medianStepsPerDay`



[C]What is the average daily pattern?
==================================================

```{r}
##Covert the data type of the date variable from "Factor" to "Date"
subActivityData$MeasureDate<-as.Date(subActivityData$MeasureDate,"%Y-%m-%d")

##Calculate the average number of steps per interval across all the days
stepsPerInterval<-with(subActivityData,tapply(Steps,Interval,mean,na.rm=TRUE))
stepsPerInterval<-data.frame(unique(subActivityData$Interval),stepsPerInterval)
stepsPerInterval<-setNames(stepsPerInterval,c("Interval","Average_Steps"))
```

C1. Plot of 5-min intervals and average number of steps taken across all days


```{r}
with(stepsPerInterval,plot(Interval,Average_Steps,type="l"))

##2 Calculate and report the Interval with max avg steps
maxstepInterval<-stepsPerInterval[which.max(stepsPerInterval$Average_Steps),1]

```

C2. The interval with the maximum number of steps is `r maxstepInterval`


[D] Imputing missing values
================================

```{r}

##1.Calculate the number of record where no data is available for Steps 
numOfNArecords<-sum(is.na(activityData$Steps))

```

D1. `r numOfNArecords` records don't have data associated with it (Steps=NA) 
 

D2. We will fill the NA values in the dataset with the mean number of steps corresponding 
to that interval. 

```{r}
##D3.Create a duplicate of the activiy data and then replace the NA value in Steps
##with the average number of steps for the corresponding interval

newActivityData<-activityData

for (i in seq_along(newActivityData$Steps)){
        
        if (is.na(newActivityData$Steps[i])){
                interval<-newActivityData$Interval[i]
                newActivityData$Steps[i]<-stepsPerInterval[stepsPerInterval$Interval==interval,2]
        }
}

## Calculate the total steps take each day with the newActivityData
newStepsPerDay<-tapply(newActivityData$Steps,newActivityData$MeasureDate,sum)
newStepsPerDay<-data.frame(unique(newActivityData$MeasureDate),newStepsPerDay)
newStepsPerDay<-setNames(newStepsPerDay,c("MeasureDate","Steps"))
```

D4.1 Histogram of total steps taken each day with imputed dataset

```{r}
hist(newStepsPerDay$Steps,xlab="Steps Per Day",main="Frequency of Steps per Day - Imputed dataset")

##4.3 Calculate and report the mean number of steps take each day with new dataset
newMeanStepsPerDay<-mean(newStepsPerDay$Steps,na.rm = TRUE)

##4.4 Calculate and report the median number of steps take each day with new dataset
newMedianStepsPerDay<-median(newStepsPerDay$Steps,na.rm = TRUE)

diffMean<-abs(newMeanStepsPerDay-meanStepsPerDay)
diffMedian<-abs(newMedianStepsPerDay-medianStepsPerDay)
```

D4.2 Steps per Day:

- Mean of all the Steps per Day with imputed dataset is: `r meanStepsPerDay`
- Median of all the Steps per Day with imputed dataset is: `r medianStepsPerDay`


D4.3 Impact of imputing the missing data on the estimates of steps per day is as follows:

- Impact on Mean: `r diffMean`
- Impact on Median: `r diffMedian`

```{r}

##Add a new variable to the newActivityData which shows day of the week for the date
newActivityData<-newActivityData%>%
                        mutate(Weekday=weekdays(as.Date(MeasureDate)))
                        
##Add a new variable that show wether a day is a "weekday" or a "weekend"
newActivityData<-newActivityData%>%
        mutate(Wkday=ifelse(Weekday%in%c("Saturday","Sunday"),"weekend","weekday"))

##Calculate average number of Steps for each of the intervals over "weekday" and "weekend"
avgStepsWeekDayorEnd<-aggregate(Steps~Wkday+Interval,data=newActivityData,mean)

```

D4.4 Plot of average number of steps per interval - segregated by weedays/weekends 

```{r}
##Plot Steps for intervals for weekday and weekend
qplot(x=Interval,y=Steps,data=avgStepsWeekDayorEnd,facets=.~Wkday)+geom_line()

```